---
title: "Project 4: Canadian Agri Market Insights — Notebook"
output:
  html_document:
    toc: true
    toc_depth: 2
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = TRUE, warning = FALSE)
required_pkgs <- c("DBI", "RSQLite", "readr", "dplyr")
for (p in required_pkgs) {
  if (!requireNamespace(p, quietly = TRUE)) {
    install.packages(p, repos = "https://cloud.r-project.org")
  }
  suppressPackageStartupMessages(library(p, character.only = TRUE))
}
project_dir <- getwd()
db_path <- file.path(project_dir, "FinalDB.sqlite")
if (file.exists(db_path)) file.remove(db_path)
conn <- DBI::dbConnect(RSQLite::SQLite(), db_path)
```

<div id="problem-01">

## Problem 1. Create Tables

Create the following tables in SQLite: `CROP_DATA`, `FARM_PRICES`, `DAILY_FX`, `MONTHLY_FX`.

```{r problem1}
# Explicit DDL to create empty tables
DBI::dbExecute(conn, "CREATE TABLE IF NOT EXISTS CROP_DATA (
  CD_ID INTEGER,
  YEAR TEXT,
  CROP_TYPE TEXT,
  GEO TEXT,
  SEEDED_AREA INTEGER,
  HARVESTED_AREA INTEGER,
  PRODUCTION INTEGER,
  AVG_YIELD INTEGER
)")

DBI::dbExecute(conn, "CREATE TABLE IF NOT EXISTS FARM_PRICES (
  CD_ID INTEGER,
  DATE TEXT,
  CROP_TYPE TEXT,
  GEO TEXT,
  PRICE_PRERMT REAL
)")

DBI::dbExecute(conn, "CREATE TABLE IF NOT EXISTS DAILY_FX (
  DFX_ID INTEGER,
  DATE TEXT,
  FXUSDCAD REAL
)")

DBI::dbExecute(conn, "CREATE TABLE IF NOT EXISTS MONTHLY_FX (
  DFX_ID INTEGER,
  DATE TEXT,
  FXUSDCAD REAL
)")

# Show confirmation of created tables
DBI::dbListTables(conn)
```

</div>

<div id="problem-02">

## Problem 2. Read Datasets and Load Tables

Read CSVs and load into the created tables.

```{r problem2}
crop_df <- readr::read_csv(file.path(project_dir, "Annual_Crop_Data.csv"), show_col_types = FALSE)
farm_df <- readr::read_csv(file.path(project_dir, "Monthly_Farm_Prices.csv"), show_col_types = FALSE)
daily_fx_df <- readr::read_csv(file.path(project_dir, "Daily_FX.csv"), show_col_types = FALSE)
monthly_fx_df <- readr::read_csv(file.path(project_dir, "Monthly_FX.csv"), show_col_types = FALSE)

DBI::dbWriteTable(conn, "CROP_DATA", crop_df, append = TRUE)
DBI::dbWriteTable(conn, "FARM_PRICES", farm_df, append = TRUE)
DBI::dbWriteTable(conn, "DAILY_FX", daily_fx_df, append = TRUE)
DBI::dbWriteTable(conn, "MONTHLY_FX", monthly_fx_df, append = TRUE)

# Show brief confirmation (row counts)
list(
  CROP_DATA = DBI::dbGetQuery(conn, "SELECT COUNT(*) AS rows FROM CROP_DATA"),
  FARM_PRICES = DBI::dbGetQuery(conn, "SELECT COUNT(*) AS rows FROM FARM_PRICES"),
  DAILY_FX = DBI::dbGetQuery(conn, "SELECT COUNT(*) AS rows FROM DAILY_FX"),
  MONTHLY_FX = DBI::dbGetQuery(conn, "SELECT COUNT(*) AS rows FROM MONTHLY_FX")
)
```

</div>

<div id="problem-03">

## Problem 3. Count FARM_PRICES Records

```{r problem3}
DBI::dbGetQuery(conn, "SELECT COUNT(*) AS record_count FROM FARM_PRICES;")
```

</div>

<div id="problem-04">

## Problem 4. Geographies in FARM_PRICES

```{r problem4}
DBI::dbGetQuery(conn, "SELECT DISTINCT GEO FROM FARM_PRICES ORDER BY GEO;")
```

</div>

<div id="problem-05">

## Problem 5. Rye harvested in Canada (1968)

```{r problem5}
DBI::dbGetQuery(conn, "SELECT HARVESTED_AREA FROM CROP_DATA
   WHERE CROP_TYPE = 'Rye' AND GEO = 'Canada' AND SUBSTR(YEAR,1,4) = '1968';")
```

</div>

<div id="problem-06">

## Problem 6. First 6 FARM_PRICES rows for Rye

```{r problem6}
DBI::dbGetQuery(conn, "SELECT * FROM FARM_PRICES
   WHERE CROP_TYPE = 'Rye'
   ORDER BY DATE ASC
   LIMIT 6;")
```

</div>

<div id="problem-07">

## Problem 7. Provinces that grew Barley

```{r problem7}
DBI::dbGetQuery(conn, "SELECT DISTINCT GEO FROM CROP_DATA
   WHERE CROP_TYPE = 'Barley' AND GEO <> 'Canada'
   ORDER BY GEO;")
```

</div>

<div id="problem-08">

## Problem 8. First and last FARM_PRICES dates

```{r problem8}
DBI::dbGetQuery(conn, "SELECT MIN(DATE) AS first_date, MAX(DATE) AS last_date FROM FARM_PRICES;")
```

</div>

<div id="problem-09">

## Problem 9. Crops with PRICE_PRERMT >= 350

```{r problem9}
DBI::dbGetQuery(conn, "SELECT DISTINCT CROP_TYPE FROM FARM_PRICES
   WHERE PRICE_PRERMT >= 350
   ORDER BY CROP_TYPE;")
```

</div>

<div id="problem-10">

## Problem 10. Saskatchewan crop types (year 2000) ranked by average yield

```{r problem10}
res10 <- DBI::dbGetQuery(conn, "SELECT CROP_TYPE, AVG(AVG_YIELD) AS avg_yield
   FROM CROP_DATA
   WHERE TRIM(GEO) = 'Saskatchewan' AND CAST(SUBSTR(YEAR,1,4) AS INTEGER) = 2000
   GROUP BY CROP_TYPE
   ORDER BY avg_yield DESC;")
res10
cat("\nBest performing crop (2000, Saskatchewan): ", if (nrow(res10) > 0) res10$CROP_TYPE[1] else "<none>")
```

</div>

<div id="problem-11">

## Problem 11. Average yield by crop and geography since 2000

```{r problem11}
res11 <- DBI::dbGetQuery(conn, "SELECT CROP_TYPE, GEO, AVG(AVG_YIELD) AS avg_yield
   FROM CROP_DATA
   WHERE CAST(SUBSTR(YEAR,1,4) AS INTEGER) >= 2000
   GROUP BY CROP_TYPE, GEO
   ORDER BY avg_yield DESC;")
head(res11, 12)
cat("\nTop since 2000: ", if (nrow(res11) > 0) paste(res11$CROP_TYPE[1], "-", res11$GEO[1]) else "<none>")
```

</div>

<div id="problem-12">

## Problem 12. Most recent wheat harvested area in Canada (subquery)

```{r problem12}
DBI::dbGetQuery(conn, "SELECT HARVESTED_AREA FROM CROP_DATA
   WHERE CROP_TYPE = 'Wheat' AND GEO = 'Canada' AND YEAR = (
     SELECT MAX(YEAR) FROM CROP_DATA WHERE CROP_TYPE = 'Wheat' AND GEO = 'Canada'
   );")
```

</div>

<div id="problem-13">

## Problem 13. Canola prices in Saskatchewan (CAD & USD), recent 6 months

```{r problem13}
DBI::dbGetQuery(conn, "SELECT fp.DATE,
          fp.CROP_TYPE,
          fp.GEO,
          fp.PRICE_PRERMT AS price_cad,
          (fp.PRICE_PRERMT / fx.FXUSDCAD) AS price_usd
   FROM FARM_PRICES fp, MONTHLY_FX fx
   WHERE fp.DATE = fx.DATE
     AND fp.CROP_TYPE = 'Canola'
     AND fp.GEO = 'Saskatchewan'
   ORDER BY fp.DATE DESC
   LIMIT 6;")
```

</div>

```{r teardown, include=FALSE}
DBI::dbDisconnect(conn)
```